Delayed dichromatism as a convenient tool to disentangle the effects of survival and productivity on the population dynamics in waterfowl

### 2024-05-28 Code/Draft_ruddy_duck_vital_rates.Rmd
after instatext

---
lang: en-EN
bibliography: "`r paste0('C:/Users/', Sys.info()[['user']], '/Dropbox/JabRef/Library_OFB.bib')`"
csl: https://www.zotero.org/styles/apa # https://osf.io/download/6hw3q/ 
link-citations: true
title: "Delayed dichromatism as a convenient tool to disentangle the effects of survival and productivity on the population dynamics in waterfowl"
author: 
- Adrien Tableau^[Office Français de la Biodiversité, France - adrien.tableau@ofb.gouv.fr]
- Iain Henderson^[Animal and Plant Health Agency, United Kingdom - iain.henderson@apha.gov.uk]
- Sébastien Reeber^[Société Nationale pour la Protection de la Nature, France - sebastien.reeber@snpn.fr]
- Matthieu Guillemain^[Office Français de la Biodiversité, France - matthieu.guillemain@ofb.gouv.fr]
- Jean-François Maillard^[Office Français de la Biodiversité, France - jean-francois.maillard@ofb.gouv.fr]
- Alain Caizergues^[Office Français de la Biodiversité, France - alain.caizergues@ofb.gouv.fr]
date: "`r Sys.setlocale('LC_TIME', 'English'); format(Sys.time(), '%d %B, %Y')`"
abstract: |
  Monitoring the number of individuals is by far the most popular strategy for investigating the environmental factors ruling population dynamics and for measuring the efficacy of management actions targeting population recovery, control or eradication. Monitoring numbers is unfortunately inefficient for assessing the mechanisms underlying demographic processes and more specifically to assess the extent to which population growth rate is affected by changes in adult survival rather than fluctuations in reproductive parameters. Usual methods to inform adult survival and productivity, such as capture-mark-recapture and analysis of game-hunting bag, rely on catching animals which suffers from three main drawbacks. Firstly, the selectivity of the catching methods can lead to a biased representation of the underlying population structure. Secondly, catching and releasing animals is quite invasive, which generates stress, and is thus an additional threat to often already endangered species. Thirdly, releasing may simply be impossible especially for invasive alien species that represent a threat to native taxa. In many waterfowl species, a sexual dichromatism is observed in adults whereas immatures of both sexes display a cover-up plumage similar to adult females. Using the case of alien Ruddy duck populations of Western Europe (a species displaying this pattern), we introduce a non-invasive method based on count data distinguishing plumage types that allows estimating the respective contribution of adult survival and productivity (also called recruitment rate) to population growth rate. We exemplify how, based on count surveys, changes in apparent male proportion (due to female-like immature males progressively acquiring adult male plumage) can be exploited to decompose population growth rate into adult survival and recruitment rates, greatly helping to understand the demographic impact of two different eradication strategies.

  **Keywords**: intrinsic growth rate - vital rate - reproductive success - fecundity - sex ratio - age ratio - Oxyura jamaicensis
output: 
    bookdown::pdf_document2:
      latex_engine: xelatex
      toc: false
      keep_md: false
      keep_tex: false
    bookdown::word_document2:
      toc: false
header-includes:
  - \usepackage{float} # allows to hold figure position
  - \usepackage[left]{lineno} # line numbering
---

@@

Assessing population growth rate is a first step towards a better understanding of factors underlying the dynamics of natural populations [e.g. @Niel2005]. It is also crucial for measuring the efficacy of management actions potentially undertaken to help population recovery, control, or eradication [@Shea1998]. Among the approaches available to managers for reaching these goals, those relying on the monitoring of the number of individuals (counts) are by far the most popular [@Rintala2022]. In many instances, these methods enable investigating environmental factors underlying changes in population size, and hence help implementing mitigating actions [@Faillettaz2019]. Unfortunately, relying on the monitoring of population abundance alone generally hinders basic understanding of the demographic mechanisms underlying changes in population growth rates. More specifically, based on counts only, it is nearly impossible to assess the extent to which population growth rate is affected by changes in adult survival rather than to variations in productivity [@Austin2000], which is defined as the product of all reproductive parameters. To assess the relative sensitivity of population growth rate to factors affecting adult survival or reproductive parameters, demographers most often rely on the monitoring of individuals by capture-mark-recapture [CMR, @Lebreton1992]. In practice, however, one cannot always rely on the capture and release of individuals owing to the legal status of the species that precludes any additional disturbance (e.g. critically endangered species) or forbid the release of alive individuals (e.g. invasive species or pests). Furthermore, although highly efficient for assessing demographic parameters, capture-mark-recapture methods suffer from potential drawbacks such as being invasive (and may affect behaviour and hence survival or breeding success) and hardly affordable when time and money are limiting. Genetic monitoring is a non-invasive alternative to capture-mark-recapture, but it suffers from being costly and requires large samples [@Caniglia2011].

Most of the time therefore, managers do the best of a bad job using counts as a viable option for tracking population size changes and then population growth rate, in order to assess the relevance of management actions. Distinguishing immatures from adults in counts enables estimating adult survival rate, i.e. the proportion of breeders that survived over one year, and productivity/recruitment rate, i.e. the number of immatures produced per breeder that reach sexual maturity, which are, by definition, recruited in the breeding population. This additional effort makes it possible to assess the relative influence of each of these components on population growth rate. Alternatively, assessing the age structure in hunting bags has been used to infer the role of decreasing reproductive success in population declines in a number of games species including ducks and geese but suffers from intractable biases [@Fox2018]. Of course, the latter approach is not suited for protected/endangered species.

Here we exemplify how, in dimorphic species, delayed maturity of males can be used for estimating adult survival and recruitment rate by differentiating male-like from female-like individuals in repeated winter counts. We used two non-native Ruddy duck (*Oxyura jamaicensis*) populations introduced in Europe as a study model. As numerous duck species, Ruddy duck is dimorphic, with newly born males looking like females until the prenuptial moult, which occurs in late winter. As a result of this late prenuptial moult of immatures, the apparent proportion of males increases over the course of the wintering season. These changes in apparent proportions of males during this period are therefore directly related to the proportion of immatures into the populations and thus to the reproductive success of the previous breeding season. We estimated adult survival and recruitment rates and assessed the impact of two different eradication strategies deployed in Great Britain and in France, respectively.

@@

The Ruddy duck is a stiff-tailed duck native from the Americas. From seven individuals initially acclimated in the Slimbridge Wetland Centre in the 1940's [@GutierrezExposito2020], a feral population began to establish in Great Britain with the first observed reproductive attempts in the wild in the 1960's (Figure \@ref(fig:count)). This feral population rapidly grew and spread to the entire country to reach more than 5,000 individuals in the early 2000s. By the end of the 1980's, a feral population began to establish also on the continent, especially in France, supposedly owing to the arrival of individuals born in Great Britain. But contrary to what was observed in Great Britain, the Ruddy ducks did not spread much in France, with the vast majority of observations and breeding attempts clustering in the North-West of the territory. In France, during winter, almost no Ruddy duck is observed outside the lake of Grand Lieu (47.09°N, 1.67°W), which greatly facilitates the monitoring of this population. Ruddy ducks are considered as a major threat to the endangered native White-headed duck (*Oxyura leucocephala*) of the South-Western Mediterranean, because of hybridization and thus an elevated risk for genetic pollution and genetic assimilation of the latter by the former [@MunozFuentes2007]. In order to mitigate the risks of genetic pollution of the White-headed duck by Ruddy ducks, eradication measures were implemented in the late 1990's both in Great Britain and in France [@GutierrezExposito2020], and were followed by a European Ruddy duck eradication plan in 1999 [@Hughes1999] (Figure \@ref(fig:count)).

The trends and status of populations of both White-headed duck and Ruddy duck are being assessed through population censuses only. It is thus impossible to assess the factors that impair recovery of the former and the efficacy of the eradication program for the latter. In particular, based on censuses alone, one cannot assess the relative effects of changes in adult survival and recruitment rates on population growth rates which are prerequisites to identifying limiting factors.

@@

Like many other ducks, Ruddy ducks and White-headed ducks both display delayed dichromatism thereby young males acquire the typical colourful plumage of their kind, and hence can be distinguished from females, over the course of the interbreeding season, typically by mid-winter for the earliest individuals [@Baldassarre2014]. Delayed dichromatism generally explains the discrepancies in the estimates of apparent proportions of males counted in winter vs counted in spring, or vs males determined from individuals culled in winter (Figures \@ref(fig:flock) & \@ref(fig:pmal)).

Interestingly, delayed dichromatism enables estimating adult survival and recruitment/productivity by monitoring the seasonal evolution of apparent sex ratio and by making few and reasonable assumptions (see next section).

@@

We monitored the demography of Ruddy duck populations in both Great Britain (GB) and France (FR) using exhaustive counts on the wintering grounds during the period between December 1 and January 31. In addition, apparent sex ratios in winter could be accessible from counts that distinguished female-like from male-like individuals in winter. Such counts were performed from 2006 to 2012 in Great Britain, which corresponded to a period of strong decline of the population (Figure \@ref(fig:count)). In France, a monitoring accounting for apparent sex was performed in 1999, 2001-2009, and 2012-2019. The French population observed a rapid growth during the first years of the monitoring and then stabilized from 2006 onwards as a result of a high culling pressure. In both countries, shooting was performed under the strict control of the authorities, meaning that the exact number of individuals culled was known. Therefore, despite the age ratio in birds shot was not ascertained before 2009 in France, it is possible to grossly qualify the culling pressure in three categories for both countries: "no culling" before 1999, "low culling" between 1999 and 2005, and "high culling" from 2006 (see Figure \@ref(fig:sup) in Supplementary Material).

@@

The proportion of immature males is the difference between the proportion of female-like individuals between winter and spring. The proportion of all immature individuals is twice this value if we assume an even sex ratio at birth [@Bellrose1961; @Blums1996] and comparable mortality rates between immature females and immature males [@Ramula2018; @Wood2021]. However, counts in spring suffer from inextricable issues, such as a differential detectability between females and males. Fortunately, the proportion of immatures can also be estimated only from winter counts by making the assumption of a known and unchanging adult sex ratio over the monitoring period [@Devineau2010].

The adult sex ratio is a prerequiste to estimate the proportion of immatures. No counts differentiating males from females were available during the breeding period to estimate adult sex ratios, but we determined the sex of each adult that was shot as part of the eradication programme both in Great Britain and in France. Nevertheless, the small number of culled individuals in France prevented from getting robust adult sex ratios. In Great Britain, the adult sex ratio estimated from the analysis of birds shot did not differ among months, so data from adults collected over the whole biological cycle were pooled to estimate adult sex ratios. The proportion of males did not statistically differ between the different years with more than 500 adults shot. This is consistent with previous findings showing that adult sex ratios in ducks are usually highly stable over the short-term [@Wood2021] but may vary over the long-term. Therefore, we pooled data from all adults shot as part as the control programme to estimate the adult sex ratio, see Equation \@ref(eq:pm).

Assuming the additive property of the binomial distribution, the proportion of immatures in a population can be deduced from the proportion of males in adults (alternatively the adult sex ratio) and the proportion of male-like individuals in the different count surveys in winter (assuming all these individuals are adult males), see Equation \@ref(eq:pic). By combining the population growth rate, which is derived from the year-to-year variation of an index of abundance of population size, to the proportion of immatures, one can derive adult survival and recruitment rates, see Equation \@ref(eq:vr) and Figure \@ref(fig:outline). A key concept is that if the absolute value of population size is known, numbers of adults and recruits can be estimated, see Equation \@ref(eq:VR).

| Variable name         | Description                                                                                                |
|:-|:-------|
| *Observation (input)* |                                                                                                            |
| $AM$                  | Total number of adult males shot as part of the eradication programme                                      |
| $AF$                  | Total number of adult females shot as part of the eradication programme                                    |
| $CML_{i, t}$          | Number of male-like individuals counted in population $i$ in year $t$                                      |
| $C_{i, t}$            | Number of individuals counted in population $i$ in year $t$                                                |
| $N_{i, t}$            | Size of the population $i$ in year $t$ (maximum number of individuals counted in the wintering population) |
| *Parameter (output)*  |                                                                                                            |
| $p(m|a)$              | Proportion of males in adults (or probability of being a male knowing it is an adult)                      |
| $p(a \cap m)_{i, t}$  | Proportion of adult males in population $i$ in year $t$                                                    |
| $p(a)_{i, t}$         | Proportion of adults in population $i$ in year $t$                                                         |
| $p(re)_{i, t}$        | Proportion of immatures/recruits in population $i$ in year $t$                                             |
| $s_{i, t}$            | Adult survival rate, i.e. proportion of individuals in year $t-1$ still alive in year $t$                  |
| $r_{i, t}$            | Recruitment rate, i.e. number of recruits in population $i$ in year $t$ per individual in year $t - 1$     |
| $S_{i, t}$            | Number of adults in population $i$ in year $t$                                                             |
| $R_{i, t}$            | Number of recruits in population $i$ in year $t$                                                           |

@@

## Validating the method {#VAL}

Nevertheless, we developed an indirect approach to estimate the maximum expected recruitment rate without exploitation and assuming that the observed density levels did not affect components of the recruitment process. Estimated values above that maximum recruitment rate would reflect failure in detection of adult males. On the other hand, null or negative recruitment rates would reflect either that immature males had acquired the male plumage before the survey, or violation of the assumption about the stability of the adult sex ratio.

Practically, maximum recruitment rate is estimated from the difference between maximum growth rate and maximum adult survival rate, see Equation \@ref(eq:GR). This relationship comes from a simple reasoning for a closed population: population size in year $t$ is equal to the number of adults that survived over year $t - 1$ plus the offspring produced in year $t - 1$ that survived until the reproduction period of year $t$, i.e. the recruitment in year $t$ [@Flint2015]. For an open population, adult survival and recruitment rates are confounded to adult and recruit migrations, respectively, but this does not corrupt the equation. This relationship becomes more complex if a species with delayed maturity is considered, see @Robertson2008.

@@

```{=tex}
\begin{align} 
N_{t} & = N_{t-1} - D_{t-1} + R_{t} \notag \\
N_{t} & = N_{t-1} - (1 - s_{t}).N_{t-1} + r_{t}.N_{t-1} \notag \\
N_{t} & = s_{t}.N_{t-1} + r_{t}.N_{t-1} \notag \\
\frac{N_{t}}{N_{t-1}} & = s_{t} + r_{t} \notag \\
\lambda_{t} & = s_{t} + r_{t} \notag \\
r_{t} & = \lambda_{t} - s_{t}  
(\#eq:GR)
\end{align}
```
Maximum adult survival was defined as a uniform distribution on the $[0.7;1]$ interval. The lower limit corresponded to the upper range of survival rates observed in waterfowl species of similar weight, and the upper limit was set to 1 because survival rates of long-lived waterfowl species are very high [@Nichols1997; @Krementz1997; @Buxton2004; @Nichols1992; @Koons2014].

In both populations, the maximum growth rates occurred during their geographical expansion phase before shooting operations (Figure \@ref(fig:count)). To estimate robust maximum population growth rates for both populations, we smoothed annual growth rate values over a consistent time-period by using a linear regression on the logarithm scale, see Equation \@ref(eq:L). We discarded Great Britain data before 1972 as the estimates were probably noisy when the population was low (below 50 indiviudals) (Figure \@ref(fig:count)). After reaching the threshold of 1,000 individuals, the Great Britain (GB) population growth showed a strong inflexion despite culling had not begun (Figure \@ref(fig:count)). This observation suggests that a negative density-dependence process might occur beyond 1,000 individuals and led to consider only the first sequence of the time series to infer on the maximum growth rate in Great Britain, i.e. 1972-1981. For the French (FR) population, the sequence without culling pressure spread over the period 1994-1999.

@@

```{=tex}
\begin{equation}
log(N_{i, t}) \sim {\sf Norm}(N0_{i, J} + log(\lambda_{i, J}).t, \sigma_{i, J})
(\#eq:L)
\end{equation}
```
Estimating vital rates within an acceptable range, as defined above, would not prove that out modelling approach correctly tracked their year-to-year variability. In order to ascertain this aspect, we check the correlation between the proportions of immatures and both survival and recruitment rates derived from counts, to estimates derived from cullings, i.e. from individuals shot as part of the eradication programme. As we restricted the dataset to years with more than 100 individuals shot in winter for sake of precision, the analysis covered only five years of the time series of the GB population. Proportions of immatures were estimated from culled individuals by checking the presence of the bursa of Fabricius, which is present in immature individuals only [@Hochbaum1942]; see Equation \@ref(eq:pis). Adult survival and recruitment rates were then estimated by combining the proportion of immatures and Equation \@ref(eq:vr).

@@

```{=tex}
\begin{equation}
SI_{i, t} \sim {\sf Binom}(p(re)_{i, t}, S_{i, t})
(\#eq:pis)
\end{equation}
```
## Assessing the impact of culling strategies {#HARV}

During period when culling pressure was high (i.e. from 2006 onwards), eradication strategies differed between Great Britain and France. In Great Britain, culling mostly occurred in winter (53.4% of adults were shot before the 30th May), whereas it was mostly done during the breeding period in France (81.6% of adults were shot after the 30th May). We assessed whether or not these strategies had a different impact on populations by comparing growth rates and the relative contributions of both recruitment and survival on them (by comparing the average values over the high culling pressure period to proxies of maximum vital rates estimated when both populations reached their maximum growth, see Section \@ref(VAL)). In France, a LIFE project was implemented by early 2019 to intensify the culling pressure, especially during winter and spring. We therefore excluded the FR time series from 2019 onward in order to provide a comparison of homogeneous culling strategies.

@@

The method successfully provided estimates and associated uncertainty for the proportion of immatures in both wintering populations. Similarly, adult survival and recruitment rates, i.e. the two components accounting for population growth rate, were also properly estimated (Figure \@ref(fig:vital)).

@@

(ref:count) Temporal changes in the number of Ruddy ducks (log scale) in Great Britain (GB) and France (FR) from 1960 to 2019, with indications of periods when apparent sex ratios were also recorded

>>>>

---
lang: en-EN
bibliography: "`r paste0('C:/Users/', Sys.info()[['user']], '/Dropbox/JabRef/Library_OFB.bib')`"
csl: https://www.zotero.org/styles/apa # https://osf.io/download/6hw3q/ 
link-citations: true
title: "Delayed dichromatism as a convenient tool to disentangle the effects of survival and productivity on the population dynamics in waterfowl"
author: 
- Adrien Tableau^[Office Français de la Biodiversité, France - adrien.tableau@ofb.gouv.fr]
- Iain Henderson^[Animal and Plant Health Agency, United Kingdom - iain.henderson@apha.gov.uk]
- Sébastien Reeber^[Société Nationale pour la Protection de la Nature, France - sebastien.reeber@snpn.fr]
- Matthieu Guillemain^[Office Français de la Biodiversité, France - matthieu.guillemain@ofb.gouv.fr]
- Jean-François Maillard^[Office Français de la Biodiversité, France - jean-francois.maillard@ofb.gouv.fr]
- Alain Caizergues^[Office Français de la Biodiversité, France - alain.caizergues@ofb.gouv.fr]
date: "`r Sys.setlocale('LC_TIME', 'English'); format(Sys.time(), '%d %B, %Y')`"
abstract: |
  Monitoring the number of individuals is by far the most popular strategy for studying the environmental factors that determine population dynamics and for measuring the effectiveness of management actions aimed at population recovery, control, or eradication. unfortunately, monitoring numbers is inefficient in identifying the mechanisms underlying demographic processes and, in particular, in assessing the extent to which population growth rate is influenced by changes in adult survival rather than variations in reproductive parameters. The usual methods for determining adult survival and productivity, such as analysing capture-mark-recapture data and game-hunting bag data, rely on capturing animals, which has three major drawbacks. Firstly, the selectivity of capture methods can lead to a distorted representation of the underlying population structure. Secondly, the capture and release of animals is quite invasive, causing stress and posing an additional threat to the often already endangered species. Thirdly, release may simply be impossible, especially for invasive alien species that represent a threat to native taxa. In many waterfowl species, sexual dichromatism is observed in adults, while immatures of both sexes display a cover-up plumage similar to that of adult females. Using two non-native populations of Ruddy duck, a species that exhibits this pattern, we present a non-invasive method to estimate the respective contribution of adult survival and productivity (also called recruitment rate) to the population growth rate. We exemplify how changes in the apparent proportion of males in counts (due to female-like immature males gradually acquiring adult male plumage) can be exploited to split the population growth rate into adult survival and recruitment rates, which contributes significantly to understanding the demographic impact of two different eradication strategies.

  **Keywords**: intrinsic growth rate - vital rate - reproductive success - fecundity - sex ratio - age ratio - Oxyura jamaicensis - harvesting
output: 
    bookdown::pdf_document2:
      latex_engine: xelatex
      toc: false
      keep_md: false
      keep_tex: false
    bookdown::word_document2:
      toc: false
header-includes:
  - \usepackage{float} # allows to hold figure position
  - \usepackage[left]{lineno} # line numbering
---

@@

Assessing the growth rate of populations is a first step towards a better understanding of the factors underlying the dynamics of natural populations [e.g. @Niel2005]. It is also crucial for measuring the effectiveness of management actions that are taken to recover, control, or eradicate populations [@Shea1998]. Among the approaches available to managers to achieve these goals, those that rely on monitoring the number of individuals (counts) are by far the most popular [@Rintala2022]. In many cases, these methods allow the investigation of environmental factors underlying changes in population size, and thus help in the implementation of mitigating actions [@Faillettaz2019]. Unfortunately, relying solely on the monitoring of population abundance generally hinders the fundamental understanding of the demographic mechanisms underlying changes in population growth rates. More specifically, on the basis of counts alone, it is almost impossible to assess the extent to which population growth rate is affected by changes in adult survival rather than to variations in productivity [@Austin2000], which is defined as the product of all reproductive parameters. To assess the relative sensitivity of population growth rate to factors affecting adult survival or reproductive parameters, demographers usually rely on the monitoring of individuals by capture-mark-recapture [@Lebreton1992]. In practise, however, capture and release of individuals cannot always be relied upon, as the legal status of the species precludes any additional disturbance (e.g. critically endangered species) or prohibits the release of live individuals (e.g. invasive species or pests). Even though capture-mark-recapture methods are efficient for assessing demographic parameters, they have potential drawbacks, such as their invasiveness, which can affect behaviour and thus survival or reproductive success, and their affordability when time and money are scarce. Genetic monitoring is a non-invasive alternative to capture-mark-recapture, but it is costly and requires large samples [@Caniglia2011].

Most of the time, therefore, managers make the best of a bad job by using counts as a viable option for tracking changes in population size and then population growth rate to assess the relevance of management actions. Distinguishing between immatures and adults in counts allows estimation of adult survival rate, i.e. the proportion of breeders that have survived for one year, and productivity/recruitment rate, i.e. the number of immatures produced per breeder that reach sexual maturity, which are, by definition, recruited in the breeding population. This additional effort makes it possible to assess the relative influence of each of these components on the population growth rate. Alternatively, the assessment of age structure in hunting bags has been used to infer the role of decreasing reproductive success in population declines in a number of game species, including ducks and geese, but suffers from intractable biases [@Fox2018]. Of course, the latter approach is not appropriate for protected/endangered species.

Here we exemplify how, in dimorphic species, delayed sexual maturity of males can be used to estimate adult survival and recruitment rates by distinguishing male-like and female-like individuals in repeated winter counts. We used two non-native European populations of Ruddy duck (*Oxyura jamaicensis*) as a study model. Like many other duck species, Ruddy duck is dimorphic: newborn males look like females until the prenuptial moult, which takes place in late winter. As a result of the late prenuptial moult of immatures, the apparent proportion of males increases over the course of the wintering season. These changes in the apparent proportion of males during this period are therefore directly related to the proportion of immatures in the populations and thus to the reproductive success of the previous breeding season. The "apparent sex ratio" method developed in this study was used to estimate adult survival and recruitment rates and evaluate the effects of two different eradication strategies used in Great Britain and France, respectively.

@@

The Ruddy duck is a stiff-tailed duck native to the American continent. Starting from seven individuals acclimatised at the Slimbridge Wetland Centre in the 1940s [@GutierrezExposito2020], a feral population began to establish in Great Britain and the first attempts at reproduction in the wild were observed in the 1960s (Figure \@ref(fig:count)). This feral population grew rapidly and spread across the country until it reached more than 5,000 individuals in the early 2000s. In the late 1980s, a feral population also began to establish on the continent, particularly in France, supposedly due to the arrival of British-born individuals. However, in contrast to observations in Great Britain, Ruddy ducks have not spread very much in France. Most observations and breeding attempts have been concentrated in the north-west of the country. In France, almost no Ruddy ducks were observed in winter outside the lake of Grand Lieu (47.09°N, 1.67°W), which greatly facilitated the monitoring of this population. Ruddy ducks are considered a major threat to the endangered native White-headed duck (*Oxyura leucocephala*) in the south-western Mediterranean, as they hybridise and thus pose an increased risk of genetic pollution and genetic assimilation of the latter by the former [@MunozFuentes2007]. To reduce the risk of genetic pollution of the White-headed duck by the Ruddy duck, eradication measures were implemented in both Great Britain and France in the late 1990s [@GutierrezExposito2020], followed by a European Ruddy duck eradication plan in 1999 [@Hughes1999] (Figure \@ref(fig:count)).

The trends and status of the White-headed duck and Ruddy duck populations used to be determined exclusively by counts. It was therefore not possible to assess the factors impairing the recovery of the former and the effectiveness of the eradication programme for the latter. In particular, counts alone cannot be used to assess the relative impact of changes in adult survival and recruitment rates on population growth rates, which is a prerequisite for identifying limiting factors.

@@

Like many other duck species, both White-headed duck and Ruddy duck exhibit delayed dichromatism, i.e. young males acquire the colourful plumage typical of their species and can therefore be distinguished from females during the pre-breeding period, typically from mid-winter in the earliest individuals [@Baldassarre2014]. Delayed dichromatism generally explains the discrepancies in estimates of apparent proportions of males counted in winter compared to proportions of males counted in spring or males identified from individuals culled in winter (Figures \@ref(fig:flock) & \@ref(fig:pmal)).

Interestingly, delayed dichromatism allows the estimation of adult survival and recruitment/productivity by observing the seasonal evolution of the apparent sex ratio and making few and reasonable assumptions (see next section).

@@

We monitored the demography of Ruddy duck populations in both Great Britain (GB) and France (FR) by conducting extensive counts on the wintering grounds between 1 December and 31 January. In addition, the apparent sex ratio could be determined using winter counts that distinguished between female-like and male-like individuals. Such counts were carried out in Great Britain from 2006 to 2012, which corresponded to a period of sharp population decline (Figure \@ref(fig:count)). In France, count surveys taking into account apparent sex were conducted in 1999, 2001-2009, and 2012-2019. The French population grew rapidly in the first years of monitoring and then stabilised from 2006 onwards as a result of high culling pressure. In both countries, culling was carried out under strict official control so that the exact number of individuals culled was known. Although the age ratio of birds culled before 2009 could not be determined with certainty in France, it was still possible to roughly categorise the culling pressure in both countries into three categories: "no culling" before 1999, "low culling" between 1999 and 2005 and "high culling" from 2006 onwards (see Figure \@ref(fig:sup) in Supplementary Material).

@@

The proportion of immature males is the difference between the proportion of female-like individuals between winter and spring. The proportion of all immature individuals is twice as high if we assume an even sex ratio at birth [@Bellrose1961; @Blums1996] and comparable mortality rates between immature females and immature males [@Ramula2018; @Wood2021]. However, spring counts suffer from intractable problems, such as the different detectability of females and males during this season. Fortunately, the proportion of immatures can also be estimated from winter counts alone, assuming a known and unchanged adult sex ratio along the monitoring period [@Devineau2010].

The adult sex ratio is therefore a prerequisite for estimating the proportion of immatures. This ratio could be derived from culling data as we determined the sex of each adult culled as part of the eradication programmes in both Great Britain and France. Nevertheless, the low number of individuals culled in France prevented from getting robust adult sex ratio. In Great Britain, adult sex ratios did not differ from month to month, so data from culled adults collected throughout the whole biological cycle were pooled together each year. The proportion of males also did not differ statistically between years in which more than 500 adults were culled. This is consistent with previous findings showing that the adult sex ratio in ducks is usually very stable in the short term [@Wood2021] but can fluctuate in the long term. Therefore, we pooled the data from all adults culled in the control programme to estimate the adult sex ratio, see Equation \@ref(eq:pm).

Assuming the additive property of the binomial distribution, the proportion of immatures in a population can be derived from the proportion of males among adults (alternatively the adult sex ratio) and the proportion of male-like individuals in the different counts in winter (assuming that all these individuals are adult males), see Equation \@ref(eq:pic). By combining the population growth rate derived from the interannual variation of an index of abundance with the proportion of immatures, adult survival and recruitment rates can be derived, see Equation \@ref(eq:vr) and Figure \@ref(fig:outline). If the absolute value of the population size is known, the number of adults and the number of recruits can be estimated, see Equation \@ref(eq:VR). We called this approach the "apparent sex ratio" method.

| Variable name         | Description                                                                                                |
|:-|:-------|
| *Observation (input)* |                                                                                                            |
| $AM$                  | Total number of adult males culled as part of the eradication programme                                      |
| $AF$                  | Total number of adult females culled as part of the eradication programme                                    |
| $CML_{i, t}$          | Number of male-like individuals counted in population $i$ in year $t$                                      |
| $C_{i, t}$            | Number of individuals counted in population $i$ in year $t$                                                |
| $N_{i, t}$            | Size of the population $i$ in year $t$ (maximum number of individuals counted in the wintering population) |
| *Parameter (output)*  |                                                                                                            |
| $p(m|a)$              | Proportion of males in adults (or probability of being a male knowing it is an adult)                      |
| $p(a \cap m)_{i, t}$  | Proportion of adult males in population $i$ in year $t$                                                    |
| $p(a)_{i, t}$         | Proportion of adults in population $i$ in year $t$                                                         |
| $p(re)_{i, t}$        | Proportion of immatures/recruits in population $i$ in year $t$                                             |
| $s_{i, t}$            | Adult survival rate, i.e. proportion of individuals in year $t-1$ still alive in year $t$                  |
| $r_{i, t}$            | Recruitment rate, i.e. number of recruits in population $i$ in year $t$ per individual in year $t - 1$     |
| $S_{i, t}$            | Number of adults in population $i$ in year $t$                                                             |
| $R_{i, t}$            | Number of recruits in population $i$ in year $t$                                                           |

@@

## Validating the "apparent sex ratio" method {#VAL}

The relevance of the method was assessed by testing the likelihood of the estimates of both vital rates. For the adult survival rate, we checked whether the values were contained in the interval $[0;1]$, and compared our estimates with those in the literature [e.g. @Buxton2004; @Krementz1997; @Nichols1992, @Nichols1997]. Validation of the recruitment rate was more difficult because if the lower limit is 0, the upper limit is a product of all maximum values of its individual components, i.e. nesting rate, nesting success, clutch-size, hatching success, pre-fledging survival, and fledging survival. These reproductive parameters may be known for some duck species, but the last one is never mentioned [e.g. @Baldassarre2014].

To overcome this problem, we developed an indirect approach to estimate the maximum expected recruitment rate without exploitation and assuming that the level of population density does not affect the components of the recruitment process. Estimated values above this maximum recruitment rate would reflect a partial detection of adult males. On the other hand, null or negative recruitment rates would indicate that immature males acquired male plumage prior to the survey, or that the assumption about the stability of the adult sex ratio was violated.

In practise, the maximum recruitment rate was estimated from the difference between the maximum growth rate and the maximum survival rate of adults, see Equation \@ref(eq:GR). This relationship resulted from a simple consideration for a closed population: the population size in year $t$ is equal to the number of adults that survived the whole year $t - 1$ plus the offspring produced in year $t - 1$ that survived to the reproductive period of year $t$, i.e. the recruitment in year $t$ [@Flint2015]. In an open population, adult survival and recruitment rates are confounded with adult and recruit migrations, respectively, but this does not change the equation. This relationship becomes more complex when a species with delayed maturity is considered, see @Robertson2008.

@@

```{=tex}
\begin{align} 
N_{t} & = N_{t-1} - D_{t-1} + R_{t} \notag \\
N_{t} & = N_{t-1} - (1 - s_{t}).N_{t-1} + r_{t}.N_{t-1} \notag \\
N_{t} & = s_{t}.N_{t-1} + r_{t}.N_{t-1} \notag \\
\frac{N_{t}}{N_{t-1}} & = s_{t} + r_{t} \notag \\
\lambda_{t} & = s_{t} + r_{t} \notag \\
r_{t} & = \lambda_{t} - s_{t}  
(\#eq:GR)
\end{align}
```
The maximum adult survival rate was defined as a uniform distribution on the interval $[0.7;1]$. The lower limit corresponded to the upper range of survival rates observed in waterfowl species of similar weight, and the upper limit was set to 1, as survival rates of long-lived waterfowl species are very high [@Nichols1997; @Krementz1997; @Buxton2004; @Nichols1992; @Koons2014].

For both populations, the maximum growth rates occurred during their geographic expansion phase, which preceded the start of the eradication programmes (Figure \@ref(fig:count)). To estimate robust maximum population growth rates for both populations, we smoothed the annual growth rates over a consistent time period by using a linear regression on the logarithm scale, see Equation \@ref(eq:L). We discarded the data for Great Britain before 1972 as the estimates were likely noisy when the population was low (below 50 individuals) (Figure \@ref(fig:count)). After reaching the threshold of 1,000 individuals, the population growth in Great Britain (GB) showed a strong inflexion, although culling had not yet started (Figure \@ref(fig:count)). This observation suggests that a negative density-dependent process could occur beyond 1,000 individuals and led us to consider only the first sequence of the time series to infer the maximum growth rate in Great Britain, i.e. 1972-1981. For the French (FR) population, the sequence without culling pressure covered the period 1994-1999.

@@

```{=tex}
\begin{equation}
log(N_{i, t}) \sim {\sf Norm}(N0_{i, J} + log(\lambda_{i, J}).t, \sigma_{i, J})
(\#eq:L)
\end{equation}
```
Estimating vital rates within an acceptable range, as defined above, would not prove that our modelling approach correctly reflects the actual interannual variability. To determine this, we compared the proportions of immatures and the vital rates derived from counts with those derived from culling, i.e. from individuals culled as part of the eradication programme. As we restricted the dataset to years with more than 100 individuals culled in winter for sake of precision, the analysis only covered five years of the GB population time series. The proportions of immatures were estimated from the culled individuals by checking the presence of the bursa of Fabricius, which is only present in immature individuals [@Hochbaum1942]; see Equation \@ref(eq:pis). Adult survival and recruitment rates were then derived from the proportion of immatures and the Equation \@ref(eq:vr).

@@

```{=tex}
\begin{equation}
SI_{i, t} \sim {\sf Binom}(p(re)_{i, t}, S_{i, t})
(\#eq:pis)
\end{equation}
```
## Assessing the effects of culling strategies {#HARV}

During the period when culling pressure was high (i.e. from 2006 onwards), eradication strategies in Great Britain and France differed. In Great Britain, most individuals were culled in winter (53.4% of adults were culled before 30 May), whereas in France, most individuals were culled during the breeding season (81.6% of adults were culled after 30 May). We investigated whether these strategies had different effects on the populations by comparing the resulting growth rates and the relative contributions of both vital rates to them (by comparing the average values during the period of high culling pressure with proxies for the maximum vital rates estimated when both populations reached their maximum growth, see Section \@ref(VAL)). In France, a LIFE project was implemented from 2019 to intensify the culling pressure, especially in winter and spring. We therefore excluded the FR time series from 2019 onwards in order to compare homogeneous culling strategies.

@@

The "apparent sex ratio" method successfully provided estimates and associated uncertainties for the proportion of immatures in both wintering populations. Adult survival and recruitment rates, i.e. the two components that determine the population growth rate, were also correctly estimated (Figure \@ref(fig:vital)).

@@

(ref:count) Temporal changes in the number of Ruddy ducks (logarithmic scale) in Great Britain (GB) and France (FR) from 1960 to 2019, with references to periods when apparent sex ratios were reported

(ref:flock) A flock of Ruddy ducks observed in winter, including ten female-like individuals with whitish striped cheek (some of which are immature males), four male-like individuals with white cheek and black cap (all adult males), three unidentified individuals (1^st^, 4^th^, and 10^th^ from left) © Jay McGowan - 3 February 2013 - Tompkins, New York, United States
